---
title: "APM466 Assignment 1"
author: "Yuexin Liu"
date: '2023-02-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Import data
```{r, include = FALSE}
library(readxl)
data <- read_excel("APM466A1.xlsx")
data <- as.data.frame(data)
```


## Set up
```{r, include = FALSE}
bonddates <- c("2023-01-16", "2023-01-17", "2023-01-18", "2023-01-19", "2023-01-20", "2023-01-23", "2023-01-24", "2023-01-25", "2023-01-26", "2023-01-27")
bonddates <- as.Date(bonddates)
maturedate <- as.Date(as.character(data$Maturity_Date), format="%m/%d/%Y")
```


## Calculate ytm
```{r, include = FALSE}
closeprices = data.frame(data$PJan16, data$PJan17, data$PJan18, data$PJan19, data$PJan20, data$PJan23, data$PJan24, data$PJan25, data$PJan26, data$PJan27)
y = rep(1:11)
ytm = data.frame(y,y,y,y,y,y,y,y,y,y)
colnames(ytm) = c("YTM Jan 16", "YTM Jan 17", "YTM Jan 18", "YTM Jan 19", "YTM Jan 20", "YTM Jan 23", "YTM Jan 24", "YTM Jan 25", "YTM Jan 26", "YTM Jan 27")

# bond maturing in Mar 23
for (i in 1:10) {
  coupon=data$Coupon[1]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[1] - bonddates[i])/365
  f <- function(x) {
    - closeprices[1,i] + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[1,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Aug 23
for (i in 1:10) {
  coupon=data$Coupon[2]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[2] - bonddates[i])/365
  f <- function(x) {
    - closeprices[2,i] + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[2,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Mar 24
for (i in 1:10) {
  coupon=data$Coupon[3]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[3] - bonddates[i])/365
  f <- function(x) {
    - closeprices[3,i] + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[3,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Sept 24
for (i in 1:10) {
  coupon=data$Coupon[4]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[4] - bonddates[i])/365
  f <- function(x) {
    - closeprices[4,i] + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[4,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Mar 25
for (i in 1:10) {
  coupon=data$Coupon[5]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[5] - bonddates[i])/365
  f <- function(x) {
    - closeprices[5,i] + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[5,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Sep 25
for (i in 1:10) {
  coupon=data$Coupon[6]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[6] - bonddates[i])/365
  f <- function(x) {
    - closeprices[6,i] + 
      coupon*exp(-x*(time2maturity - 2.5)) + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[6,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Mar 26
for (i in 1:10) {
  coupon=data$Coupon[7]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[7] - bonddates[i])/365
  f <- function(x) {
    - closeprices[7,i] + 
      coupon*exp(-x*(time2maturity - 3)) + 
      coupon*exp(-x*(time2maturity - 2.5)) + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[7,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Sep 26
for (i in 1:10) {
  coupon=data$Coupon[8]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[8] - bonddates[i])/365
  f <- function(x) {
    - closeprices[8,i] + 
      coupon*exp(-x*(time2maturity - 3.5)) + 
      coupon*exp(-x*(time2maturity - 3)) + 
      coupon*exp(-x*(time2maturity - 2.5)) + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[8,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Mar 27
for (i in 1:10) {
  coupon=data$Coupon[9]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[9] - bonddates[i])/365
  f <- function(x) {
    - closeprices[9,i] + 
      coupon*exp(-x*(time2maturity - 4)) + 
      coupon*exp(-x*(time2maturity - 3.5)) + 
      coupon*exp(-x*(time2maturity - 3)) + 
      coupon*exp(-x*(time2maturity - 2.5)) + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[9,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Sep 27
for (i in 1:10) {
  coupon=data$Coupon[10]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[10] - bonddates[i])/365
  f <- function(x) {
    - closeprices[10,i] + 
      coupon*exp(-x*(time2maturity - 4.5)) + 
      coupon*exp(-x*(time2maturity - 4)) + 
      coupon*exp(-x*(time2maturity - 3.5)) + 
      coupon*exp(-x*(time2maturity - 3)) + 
      coupon*exp(-x*(time2maturity - 2.5)) + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[10,i] = uniroot(f, interval = c(-100, 100))$root
}

# bond maturing in Mar 28
for (i in 1:10) {
  coupon=data$Coupon[11]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[11] - bonddates[i])/365
  f <- function(x) {
    - closeprices[11,i] + 
      coupon*exp(-x*(time2maturity - 5)) + 
      coupon*exp(-x*(time2maturity - 4.5)) + 
      coupon*exp(-x*(time2maturity - 4)) + 
      coupon*exp(-x*(time2maturity - 3.5)) + 
      coupon*exp(-x*(time2maturity - 3)) + 
      coupon*exp(-x*(time2maturity - 2.5)) + 
      coupon*exp(-x*(time2maturity - 2)) + 
      coupon*exp(-x*(time2maturity - 1.5)) + 
      coupon*exp(-x*(time2maturity - 1)) + 
      coupon*exp(-x*(time2maturity - 0.5)) + 
      coupon*exp(-x*(time2maturity)) + 
      face*exp(-x*(time2maturity))
  }
  ytm[11,i] = uniroot(f, interval = c(-100, 100))$root
}
```


## Interpolate to calculate the Sep 23 yield for plotting
```{r, include = FALSE}
ytmnew <- ytm
for (i in 1:10) {
  time2maturity2 = as.numeric(maturedate[2] - bonddates[i])/365
  time2maturity3 = as.numeric(maturedate[3] - bonddates[i])/365
  ytmnew[2,i] = ytmnew[3,i] -((ytmnew[3,i] - ytmnew[2,i])/(time2maturity3 - time2maturity2))*0.5
}
```


## YTM Plot
```{r, echo = FALSE, fig.height = 2.5}
library(ggplot2)
Maturity = factor(c(0.1, 0.6, 1.1, 1.6, 2.1, 2.6, 3.1, 3.6, 4.1, 4.6, 5.1))
colorsytm = c("Jan 16" = "red", "Jan 17" = "blue", "Jan 18" = "black", 
              "Jan 19" = "darkgreen", "Jan 20" = "yellow", "Jan 23" = "purple",
              "Jan 24" = "pink", "Jan 25" = "steelblue", "Jan 26" = "brown",
              "Jan 27" = "limegreen")
ggplot(ytmnew, aes(x = Maturity)) +
    geom_line(aes(y = `YTM Jan 16`, group = 1, color = "Jan 16")) +
    geom_line(aes(y = `YTM Jan 17`, group = 1, color = "Jan 17")) +
    geom_line(aes(y = `YTM Jan 18`, group = 1, color = "Jan 18")) +
    geom_line(aes(y = `YTM Jan 19`, group = 1, color = "Jan 19")) +
    geom_line(aes(y = `YTM Jan 20`, group = 1, color = "Jan 20")) +
    geom_line(aes(y = `YTM Jan 23`, group = 1, color = "Jan 23")) +
    geom_line(aes(y = `YTM Jan 24`, group = 1, color = "Jan 24")) +
    geom_line(aes(y = `YTM Jan 25`, group = 1, color = "Jan 25")) +
    geom_line(aes(y = `YTM Jan 26`, group = 1, color = "Jan 26")) +
    geom_line(aes(y = `YTM Jan 27`, group = 1, color = "Jan 27")) +
    labs(x = "Time to Maturity (Years)",
         y = "Yield to Maturity",
         color = "Date",
         title = "Yield to Maturity Plot") +
    scale_color_manual(values = colorsytm) +
  guides(color = guide_legend(override.aes = list(size = 0.5))) +
  theme(legend.key.size = unit(0.1, 'cm'), 
        legend.key.height = unit(0.1, 'cm'),
        legend.key.width = unit(0.5, 'cm'), 
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8),
        axis.title.x = element_text(size=8),
        axis.title.y = element_text(size=8),
        title = element_text(size=8)) 

ggsave(file="ytm_plot.png", width=12, height=4, dpi=300)
```


## Calculate Spot Rate
```{r, include = FALSE}
closeprices = data.frame(data$PJan16, data$PJan17, data$PJan18, data$PJan19, data$PJan20, data$PJan23, data$PJan24, data$PJan25, data$PJan26, data$PJan27)
y = rep(1:11)
spotrate = data.frame(y,y,y,y,y,y,y,y,y,y)
colnames(spotrate) = c("Spot Rate Jan 16", "Spot Rate Jan 17", "Spot Rate Jan 18", "Spot Rate Jan 19", "Spot Rate Jan 20", "Spot Rate Jan 23", "Spot Rate Jan 24", "Spot Rate Jan 25", "Spot Rate Jan 26", "Spot Rate Jan 27")

# spot rate for bond maturing in Mar 23
for (i in 1:10) {
  coupon=data$Coupon[1]*(100/2)
  face=100
  time2maturity = as.numeric(maturedate[1] - bonddates[i])/365
  spotrate[1,i]=-(log(closeprices[1,i]/(coupon+face)))/(time2maturity)
}

# spot rate for the rest of the bonds
for (j in 2:11) {
  
  for (i in 1:10) {
    price = closeprices[j,i]
    coupon = data$Coupon[j]*100/2
    PVallcoupons = 0
    time2maturity = as.numeric(maturedate[j] - bonddates[i])/365
    coupon_times = seq(time2maturity, 0, -0.5)
    
    # find the present value of the all remaining but the last coupon payments
    len = length(coupon_times)
    k = len
    while (k>1){
        PVallcoupons = PVallcoupons + coupon * exp(-spotrate[len-k+1,i] * coupon_times[k])
        k = k-1
    }
    
   # calculate spot rate
    spotrate[j,i] = -(log((price - PVallcoupons)/(coupon+face)))/
      (time2maturity)

  }
  
}

```



## Spot Rate Plot
```{r, echo = FALSE, fig.height = 2.5}
library(ggplot2)
Maturity = factor(c(0.1, 0.6, 1.1, 1.6, 2.1, 2.6, 3.1, 3.6, 4.1, 4.6, 5.1))
colorsytm = c("Jan 16" = "red", "Jan 17" = "blue", "Jan 18" = "black", 
              "Jan 19" = "darkgreen", "Jan 20" = "yellow", "Jan 23" = "purple",
              "Jan 24" = "pink", "Jan 25" = "steelblue", "Jan 26" = "brown",
              "Jan 27" = "limegreen")
ggplot(spotrate, aes(x = Maturity)) +
    geom_line(aes(y = `Spot Rate Jan 16`, group = 1, color = "Jan 16")) +
    geom_line(aes(y = `Spot Rate Jan 17`, group = 1, color = "Jan 17")) +
    geom_line(aes(y = `Spot Rate Jan 18`, group = 1, color = "Jan 18")) +
    geom_line(aes(y = `Spot Rate Jan 19`, group = 1, color = "Jan 19")) +
    geom_line(aes(y = `Spot Rate Jan 20`, group = 1, color = "Jan 20")) +
    geom_line(aes(y = `Spot Rate Jan 23`, group = 1, color = "Jan 23")) +
    geom_line(aes(y = `Spot Rate Jan 24`, group = 1, color = "Jan 24")) +
    geom_line(aes(y = `Spot Rate Jan 25`, group = 1, color = "Jan 25")) +
    geom_line(aes(y = `Spot Rate Jan 26`, group = 1, color = "Jan 26")) +
    geom_line(aes(y = `Spot Rate Jan 27`, group = 1, color = "Jan 27")) +
    labs(x = "Time to Maturity (Years)",
         y = "Spot Rate",
         color = "Date",
         title = "Spot Rate Plot") +
    scale_color_manual(values = colorsytm) +
  guides(color = guide_legend(override.aes = list(size = 0.5))) +
  theme(legend.key.size = unit(0.1, 'cm'), 
        legend.key.height = unit(0.1, 'cm'),
        legend.key.width = unit(0.5, 'cm'), 
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8),
        axis.title.x = element_text(size=8),
        axis.title.y = element_text(size=8),
        title = element_text(size=8)) 

ggsave(file="spotrate_plot.png", width=12, height=4, dpi=300)
```



## Calculate Forward Rate
```{r, echo = FALSE}
y = rep(1:4)
forward = data.frame(y,y,y,y,y,y,y,y,y,y)
colnames(forward) = c("Forward Rate Jan 16",  "Forward Rate Jan 17", "Forward Rate Jan 18", "Forward Rate Jan 19", "Forward Rate Jan 20", "Forward Rate Jan 23", "Forward Rate Jan 24", "Forward Rate Jan 25", "Forward Rate Jan 26", "Forward Rate Jan 27")

for (j in 1:4) {
  for (i in 1:10) {
    forward[j,i] = (spotrate[3+2*j,i]*(j+1) - spotrate[3,i])/j
  }
 }

```



## Forward Rate Plot
```{r, echo = FALSE, fig.height = 2.5}
library(ggplot2)
Maturity = factor(c("1yr-1yr", "1yr-2yr", "1yr-3yr", "1yr-4yr"))
colorsytm = c("Jan 16" = "red", "Jan 17" = "blue", "Jan 18" = "black", 
              "Jan 19" = "darkgreen", "Jan 20" = "yellow", "Jan 23" = "purple",
              "Jan 24" = "pink", "Jan 25" = "steelblue", "Jan 26" = "brown",
              "Jan 27" = "limegreen")
ggplot(forward, aes(x = Maturity)) +
    geom_line(aes(y = `Forward Rate Jan 16`, group = 1, color = "Jan 16")) +
    geom_line(aes(y = `Forward Rate Jan 17`, group = 1, color = "Jan 17")) +
    geom_line(aes(y = `Forward Rate Jan 18`, group = 1, color = "Jan 18")) +
    geom_line(aes(y = `Forward Rate Jan 19`, group = 1, color = "Jan 19")) +
    geom_line(aes(y = `Forward Rate Jan 20`, group = 1, color = "Jan 20")) +
    geom_line(aes(y = `Forward Rate Jan 23`, group = 1, color = "Jan 23")) +
    geom_line(aes(y = `Forward Rate Jan 24`, group = 1, color = "Jan 24")) +
    geom_line(aes(y = `Forward Rate Jan 25`, group = 1, color = "Jan 25")) +
    geom_line(aes(y = `Forward Rate Jan 26`, group = 1, color = "Jan 26")) +
    geom_line(aes(y = `Forward Rate Jan 27`, group = 1, color = "Jan 27")) +
    labs(x = "Term",
         y = "Forward Rate",
         color = "Date",
         title = "Forward Rate Plot") +
    scale_color_manual(values = colorsytm) +
  guides(color = guide_legend(override.aes = list(size = 0.5))) +
  theme(legend.key.size = unit(0.1, 'cm'), 
        legend.key.height = unit(0.1, 'cm'),
        legend.key.width = unit(0.5, 'cm'), 
        legend.title = element_text(size=8), 
        legend.text = element_text(size=8),
        axis.title.x = element_text(size=8),
        axis.title.y = element_text(size=8),
        title = element_text(size=8)) 

ggsave(file="forwardrate_plot.png", width=12, height=4, dpi=300)
```



## Covariance matrix for daily log-returns of yield
```{r, include = FALSE}
y = rep(1:9)
logreturnyield = data.frame(y,y,y,y,y)
for (j in 1:9) {
  for (i in 1:5) {
  logreturnyield[j,i]=log(ytmnew[1+2*i,j+1]/ytmnew[1+2*i,j])
  }
}

covlogreturnyield = cov(logreturnyield, logreturnyield)
covlogreturnyield
```


## Covariance matrix for log-returns of forward rate
```{r, include = FALSE}
y = rep(1:9)
logreturnforward = data.frame(y,y,y,y)
for (j in 1:9) {
  for (i in 1:4) {
  logreturnforward[j,i]=log(forward[i,j+1]/forward[i,j])
  }
}

covlogreturnforward = cov(logreturnforward, logreturnforward)
covlogreturnforward
```


## Print covariance matrices
```{r, echo=FALSE}
t1 <- covlogreturnyield
colnames(t1) <- c('$X_1$', '$X_2$', '$X_3$', '$X_4$', '$X_5$')
rownames(t1) <- c('$X_1$', '$X_2$', '$X_3$', '$X_4$', '$X_5$')
t2 <- covlogreturnforward
colnames(t2) <- c('$X_1$', '$X_2$', '$X_3$', '$X_4$')
rownames(t2) <- c('$X_1$', '$X_2$', '$X_3$', '$X_4$')
knitr::kable(t1, caption = "Covariance matrix for log returns of yield")
knitr::kable(t2, caption = "Covariance matrix for log returns of the forward rate")
```


## Calculate the eigenvalues and eigenvectors for the matrix of daily log returns of yield
```{r, include = FALSE}
eigen_yield=eigen(covlogreturnyield,symmetric=TRUE)
eigen_yield$values
eigen_yield$vectors
```


## Calculate the eigenvalues and eigenvectors for the forward rates
```{r, include = FALSE}
eigen_forward=eigen(covlogreturnforward,symmetric=TRUE)
```


## Print the eigenvalues and eigenvectors
```{r, echo = FALSE}
t3 <- data.frame(eigen_yield$values[1], eigen_yield$values[2], eigen_yield$values[3], eigen_yield$values[4], eigen_yield$values[5])
rownames(t3) <- NULL
colnames(t3) <- c("$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$", "$\\lambda_4$", "$\\lambda_5$")
t4 <- data.frame(eigen_yield$vectors)
rownames(t4) <- NULL
colnames(t4) <- c("$v_1$", "$v_2$", "$v_3$", "$v_4$", "$v_5$")
t5 <- data.frame(eigen_forward$values[1], eigen_yield$values[2], eigen_yield$values[3], eigen_yield$values[4])
rownames(t5) <- NULL
colnames(t5) <- c("$\\lambda_1$", "$\\lambda_2$", "$\\lambda_3$", "$\\lambda_4$")
t6 <- data.frame(eigen_forward$vectors)
rownames(t6) <- NULL
colnames(t6) <- c("$v_1$", "$v_2$", "$v_3$", "$v_4$")
knitr::kable(t3, caption = "Eigenvalues for covariance matrix for log returns of yield", escape = FALSE)
knitr::kable(t4, caption = "Eigenvectors for covariance matrix for log returns of yield", escape = FALSE)
knitr::kable(t5, caption = "Eigenvalues for covariance matrix for log returns of the forward rate", escape = FALSE)
knitr::kable(t6, caption = "Eigenvectors for covariance matrix for log returns of the forward rate", escape = FALSE)
```